---
title: 'dplyr::count misses factor levels: a case in comparing scRNAseq cell type
  abundance '
author: Ming Tang
date: '2020-08-26'
slug: dplyr-count-misses-factor-levels-a-case-in-comparing-scrnaseq-cell-type-abundance
categories:
  - bioinformatics
  - scRNAseq
tags:
  - scRNAseq
  - tidyverse
  - R
header:
  caption: ''
  image: ''
  preview: yes
editor_options: 
  chunk_output_type: console
---

It is very common to see in the scRNAseq papers that the authors compare cell type abundance across groups (e.g., treatment vs control, responder vs non-responder).

Let's create some dummy data.

```{r  warning=FALSE, message=FALSE}
library(tidyverse)
set.seed(23)

# we have 6 treatment samples and 6 control samples, 3 clusters A,B,C
# but in the treatment samples, cluster C is absent (0 cells) in sample7
sample_id<- c(paste0("sample", 1:6, "_control", rep(c("_A","_B","_C"),each = 6)), paste0("sample", 8:12, "_treatment", rep(c("_A","_B", "_C"), each = 5)))

sample_id<- c(sample_id, "sample7_treatment_A", "sample7_treatment_B")
cell_id<- paste0("cell", 1:20000)

cell_df<- tibble::tibble(sample_id = sample(sample_id, size = length(cell_id), replace = TRUE), 
         cell_id = cell_id) %>%
  tidyr::separate(sample_id, into = c("sample_id", "group", "cluster_id"), sep= "_")


cell_num<- cell_df %>% 
  group_by(group, cluster_id, sample_id)%>%
  summarize(n=n()) 

cell_num

total_cells<- cell_df %>%
  group_by(sample_id) %>%
  summarise(total = n())

total_cells

```

join the two dataframe to get percentage of cells per cluster per sample

```{r}
cell_percentage<- left_join(cell_num, total_cells) %>%
  mutate(percentage = n/total)

cell_percentage
```

Let's plot a boxplot

```{r}
cell_percentage %>%
  ggplot(aes(x = group, y = percentage)) +
  geom_boxplot(outlier.shape = NA, aes(fill = group)) +
  geom_jitter() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~cluster_id) +
  theme_bw()+
  xlab("")
```

**YES, if you are careful enough**, you will find the treatment group in cluster C only contains 5 points.
Because if a cluster is completely missing for a sample, there will not be any cells in the original `cell_df`. However, the percentage should be 0% for that data point and you should show it in the boxplot as the jitter point. Otherwise, the result can be misleading. You can spot on such mistakes when you plot out the points on top of the boxplot.


### How to fix it

The trick is to make the factor contains all the levels of all the combinations. and when use `group_by` or `count`, add `.drop =FALSE`.

```{r}
sample_id<- paste0("sample", 1:12)
cluster_id<- c("A","B","C")


factor_levels<- tidyr::expand_grid(sample_id, cluster_id) %>%
  mutate(group = c(rep("control", 18), rep("treatment", 18))) %>%
  mutate(sample_id = paste(sample_id, cluster_id, group, sep="_"))

cell_num2<- cell_df %>%
  mutate(sample_id = paste(sample_id, cluster_id, group, sep="_")) %>%
  mutate(sample_id = factor(sample_id, levels = factor_levels$sample_id)) %>%
  group_by(sample_id, .drop=FALSE) %>%
  summarise(n=n()) %>%
  tidyr::separate(sample_id, c("sample_id", "cluster_id", "group")) 


## the 0 correctly showed up
cell_num2 %>%
  filter(sample_id == "sample7")
```

Let's replot the boxplot and see the difference:

```{r}

cell_percentage<- left_join(cell_num2, total_cells) %>%
  mutate(percentage = n/total)


# replot the same boxplot
cell_percentage %>%
  ggplot(aes(x = group, y = percentage)) +
  geom_boxplot(outlier.shape = NA, aes(fill = group)) +
  geom_jitter() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~cluster_id) +
  theme_bw()+
  xlab("")
```

Now the 0 percentage point for sample7 in cluster C showed up.

### Conclusion

* Be careful with the 0 count cell in some clusters in some samples. If you work with `Seurat`, people tend to use `seurat_obj@meta.data %>% group_by(cluster_id, sample_id, group)`, but this will miss the samples in which some clusters are missing.

* For differential abundance comparison between treatment vs control, directly comparing percentages are not optimal. Follow tutorial on using raw cell counts https://osca.bioconductor.org/multi-sample-comparisons.html# 

* More tools can be found at https://github.com/crazyhottommy/scRNAseq-analysis-notes#cell-type-prioritizationdifferential-abundance-test

