---
title: Align multiple ggplot2 plots by axis
author: Ming Tang
date: '2020-02-07'
slug: align-multiple-ggplot2-plots-by-axis
categories:
  - rstats
  - visualization
tags:
  - rstats
header:
  caption: ''
  image: ''
---

I used to use `cowplot` to align multiple ggplot2 plots but when the x-axis are of different ranges, some extra work is needed to align the axis as well.

The other day I was reading a [blog post](https://mp.weixin.qq.com/s/V1UiR98K6vy00PwkgkTmvA) by `GuangChuang Yu` and he exactly tackled this problem. His packages such as `ChIPseeker`, `ClusterProfiler`, `ggtree` are quite popular among the users.

Some dummy example from his post:

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggstance)
library(cowplot)
# devtools::install_github("YuLab-SMU/treeio")
# devtools::install_github("YuLab-SMU/ggtree")
library(tidytree)
library(ggtree)

no_legend=theme(legend.position='none')

d <- group_by(mtcars, cyl) %>% summarize(mean=mean(disp), sd=sd(disp))
d2 <- dplyr::filter(mtcars, cyl != 8) %>% rename(var = cyl)

p1 <- ggplot(d, aes(x=cyl, y=mean)) +
    geom_col(aes(fill=factor(cyl)), width=1) +
    no_legend
p2 <- ggplot(d2, aes(var, disp)) +
    geom_jitter(aes(color=factor(var)), width=.5) +
    no_legend

p3 <- ggplot(filter(d, cyl != 4), aes(mean, cyl)) +
    geom_colh(aes(fill=factor(cyl)), width=.6) +
    coord_flip() + no_legend

pp <- list(p1, p2, p3)
plot_grid(plotlist=pp, ncol=1, align='v')
```

specifying `aling='v'` aligns the plots vertically, but because the axis limits are different the x-axis is not aligned.

Let's use `coord_cartesian` to expand the xlim without filtering out the data.

```{r}
p11<- p1 + coord_cartesian(xlim = c(3,11))
p22<- p2 + coord_cartesian(xlim = c(3,11))
p33<- p3 <- ggplot(filter(d, cyl != 4), aes(cyl, mean)) +
    geom_col(aes(fill=factor(cyl)), width=.6) +
  coord_cartesian(xlim = c(3,11)) +no_legend

pp1 <- list(p11, p22, p33)
plot_grid(plotlist=pp1, ncol=1, align='v')
```

This works.

However, as mentioned in the blog post by `GuangChuang Yu`. There are several other cases that this may not be easy to work out:

* what if the x-axis is character string rather than continuous digits?
* what if the first plot is not from a dataframe (e.g. a tree object from ggtree)

Let's use the other example from the blog post.

```{r}
set.seed(2019-10-31)
tr <- rtree(10)

d1 <- data.frame(
    # only some labels match
    label = c(tr$tip.label[sample(5, 5)], "A"),
    value = sample(1:6, 6))

d2 <- data.frame(
    label = rep(tr$tip.label, 5),
    category = rep(LETTERS[1:5], each=10),
    value = rnorm(50, 0, 3))

g <- ggtree(tr) + geom_tiplab(align=TRUE)
g
```

This is a tree. 

Make some other dummy dataframe for making a bar plot and a heatmap:

```{r warning=FALSE}
d <- filter(g, isTip) %>% select(c(label, y))

dd1 <- left_join(d1, d, by='label')
dd1
dd2 <- left_join(d2, d, by='label')
head(dd2)

# a bar graph
p1 <- ggplot(dd1, aes(y, value)) + geom_col(aes(fill=label)) +
    coord_flip() + theme_tree2() + theme(legend.position='none')
 
# a heatmap
p2 <- ggplot(dd2, aes(x=category, y=y)) +
    geom_tile(aes(fill=value)) + scale_fill_viridis_c() +
    theme_tree2() + theme(legend.position='none')

cowplot::plot_grid(g, p1, p2, ncol=3, align='h',
    labels=LETTERS[1:3], rel_widths = c(1, .5, .8))
```

The y-axis is not aligned with the tip of the `ggtree` output if you read carefully.

Let's use the `ylim2` function from `ggtree`:

```{r warning=FALSE}
p1 <- p1 + ylim2(g)
p2 <- p2 + ylim2(g)

cowplot::plot_grid(g, p1, p2, ncol=3, align='h',
    labels=LETTERS[1:3], rel_widths = c(1, .5, .8))
```

Now they are aligned perfectly!

`ggtree::ylim2()` and `ggtree::xlim2()` can be very useful for other cases. Thanks `GuangChuang Yu` for making it!