---
title: compare slopes in linear regression
author: Ming Tang
date: '2020-10-02'
slug: compare-slopes-in-linear-regression
categories:
  - statistics
  - rstats
tags:
  - statistics
  - rstats
  - linear-regresssion
header:
  caption: ''
  image: ''
  preview: yes
editor_options: 
  chunk_output_type: console
---

I asked this question on [twitter](https://twitter.com/tangming2005/status/1308225162324647937).

### load the package

```{r warning=FALSE, message=FALSE}
library(tidyverse)

```


### make some dummy data 
The dummy example: We have two groups of samples: disease and health.
We treat those cells in vitro with different dosages (0, 1, 5) of a chemical X and count the cell number after 3 hours.

```{r}
x <- tibble(
  '0' = c(8.66,
               11.50,
               7.01,
               13.40,
               11.30,
               8.13,
               5.92,
               7.54),
  '1' = c(22.10,
               23.00,
               22.00,
               35.70,
               32.60,
               26.90,
               23.30,
               21.60),
  '5' = c(31.20,
               30.60,
               28.40,
               37.30,
               41.40,
               40.20,
               29.20,
               35.40))

y <- tibble(
  '0' = c(13.90,
               5.67,
               2.59,
               9.77,
               5.63,
               4.59,
               7.35,
               5.92),
  '1' = c(12.60,
               8.48,
               5.59,
               5.43,
               13.10,
               18.00,
               23.90,
               14.50),
  '5' = c(20.00,
               10.70,
               9.73,
               16.20,
               15.10,
               21.30,
               25.60,
               18.40))

x<- x %>% tidyr::gather(1:3, key= "dose", value = "cells") %>%
  mutate(group = "Health")

y<- y %>% tidyr::gather(1:3, key= "dose", value = "cells") %>%
  mutate(group = "Disease")

z<- rbind(x,y)

ggplot(z, aes(x= dose, y = cells)) +
  geom_boxplot(aes(color = dose)) +
  geom_point() +
  facet_wrap(~group) +
  theme_bw(base_size = 14)
```


We see that as the dosage increases, the cell number increases in both disease and healthy groups.
However, the question is that if the cell number increases faster in the healthy group than the disease group or not? How can I get a p-value for that?

As [Chenxin Li](https://twitter.com/ChenxinLi2/status/1308232791709024257) pointed out. The dose has to be `numeric`:
```{r}
z<- z %>%
  mutate(dose = as.numeric(dose)) 

lm_res<- lm(cells ~  dose *group, data = z)
summary(lm_res)

```

### Interpret the result

$beta 0 = 8.7$ is the intercept.  
$beta 1 = 1.77$ is the coefficient of the `dose`. It is interpreted as the unique effect of dose when group is disease (reference group). 
$beta 2 = 6.05$ is the coefficient of the `group`. It is interpreted as the effect of group when the dose is 0. 
$beta 3 = 2.39$ is the coefficient of the interaction term `dose:group`. $beta3$ is the **difference** in the slopes of non-reference minus reference (reference is the Disease group chosen by R alphabetically).

In other words, the slope of Healthy group - Disease group = 2.39, which means the cell number in the healthy group increases faster. The p-value is 0.00626


I want to thank Collin in Shirley's lab and other tweeps for helping out for my question.

I found [this explanation](https://www.theanalysisfactor.com/interpreting-interactions-in-regression/) of how to inteprete the interaction term is very good. I highly recommend reading through this as well: https://lindeloev.github.io/tests-as-linear/#62_two-way_anova_(plot_in_progress) 


