---
title: Be careful when left_join tables with duplicated rows
author: Ming Tang
date: '2021-10-17'
slug: be-careful-when-left-join-tables-with-duplicated-rows
categories:
  - bioinformatics
  - R
tags:
  - R
  - data
header:
  caption: ''
  image: ''
  preview: yes
editor_options: 
  chunk_output_type: console
---

This is going to be a really short blog post. I recently found that if I join two tables with one of the tables having duplicated rows, the final joined table also contains the duplicated rows. It could be the expected behavior for others but I want to make a note here for myself.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

df1<- tibble(key = c("A", "B", "C", "D", "E"),
             value = 1:5)

df1
```


dataframe 2 has two identical rows for `B`.

```{r}
df2<- tibble(key = c("B", "B", "D", "C", "A", "E"),
             value2 = c(5, 5:9))

df2
```


Let's see how the `left_join` behaviors:

```{r}
left_join(df1, df2)
```

You see there are two identical rows for `B`!

If that's not what you want, you can either filter `df2` first or filter the the joined table.

```{r}
left_join(df1, df2) %>%
  distinct(key, value, value2, .keep_all =TRUE)
```

