---
title: Basic tensor/array manipulations in R
author: Ming Tang
date: '2023-03-18'
slug: basic-tensor-array-manipulations-in-r
categories:
  - machine learning
tags:
  - deeplearning
  - machine learning
header:
  caption: ''
  image: ''
editor_options: 
  chunk_output_type: console
---

### Sign up for my newsletter to not miss a post like this

https://divingintogeneticsandgenomics.ck.page/newsletter


```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message= FALSE, warning = FALSE
)
```

In my last post, I showed you how to build a neural network in `Keras` with less than 20 lines of code.
One of the key road blocks for beginners is to transform the input to the right shape of tensor (the deep learning terminology) or 
array (the R built-in type).

In this post, I am going to show you some basic manipulations of the array.

```{r}
library(keras)
set.seed(123)
```


### Construct tensors

A 2D tensor is just a matrix.

```{r}
x<- sample(c(0,1), size= 24, replace = TRUE)
x
y<- matrix(x, nrow = 4)
y
```
The matrix is filled in a row-wise manner.


A 3D tensor is a 3D array.

```{r}
y<- array(x, dim = c(2,3,4))
```

```{r}
x

y[,,1]
y[,,2]
y[,,3]
```

pay attention on how the arrays are filled for each dimension. 
If you use `array` to construct the tensor, it fills the elements starting from the
last dimension in the column wise manner.


The first dimension is usually the sample. We can think it this 3D tensor represents
two samples, and each sample has a 2D matrix to represent the time and the feature in
timeseries data.

```{r}
y[1,,]
y[2,,]
```


### reshape the array

If we use the `keras::array_reshape` function.

>by default, array_reshape() will fill the new dimensions in row-major (C-style) ordering, while dim<-() will fill new dimensions in column-major (Fortran-style) ordering. This is done to be consistent with libraries like NumPy, Keras, and TensorFlow, which default to this sort of ordering when reshaping arrays.

```{r}
# the default is the C-stype: order = "C"
z<- array_reshape(x, dim = c(2,3,4))   

x
z[1,,]
z[2,,]

```
Now, it fills in the array starting from the first dimension and in a row-wise manner.

```{r}
n<- array_reshape(x, dim = c(2,3,4), order = "F") 

x
n[,,1]
n[,,2]

```

Now, it fills in the array staring from the last dimension and in a column-wise manner.
It is the same as the `dim()` method output:

```{r}
all.equal(y, n)
```


### permutate index 

```{r}
mat <- replicate(3, matrix(runif(24),ncol=4), simplify=FALSE )

mat
```

It is a list of 3 matrices, each matrix is for one sample.

Turn it to an array:
```{r}
array1<- simplify2array(mat)
dim(array1)

# the last dimension is the sample.
array1[,,1]
```


How can I permutate it to have the first dimension to be 3?

```{r}
array2<- aperm(array1,  c(3,1,2))
dim(array2)

array2[1,,]

all.equal(mat[[1]], array2[1,,])
```

Use `unlist` to vectorize the list of matrices. `unlist` collects all the elements
in a column-wise manner.

```{r}
unlist(mat) %>% head(n = 10)
mat[[1]]

array3<- unlist(mat) %>% array(dim= c(6,4,3))
all.equal(array1, array3)
```


```{r}
array4<- unlist(mat) %>% array_reshape(dim=c(3,6,4))

array4[1,,]
mat[[1]]
```

`array_reshap` put the sample in the first dimension, but it fills in the matrix in a row-wise manner.

We can get all the  element in a row-wise manner:

```{r}
# column-wise
mat[[1]]
mat[[1]] %>% c()

# row-wise, we first transpose it 
mat[[1]] %>% t() 

mat[[1]] %>% t() %>% c()
```


```{r}
mat1<- lapply(mat, function(x) x %>%t() %>% c()) %>% unlist()

array5<- array_reshape(mat1, dim=c(3,6,4))
array5[1,,]
mat[[1]]

all.equal(array5[1,,], mat[[1]])
all.equal(array5, array2)
```

### tensor operations

In R, everything is vectorized, so you can do element-wise multiplication, subtraction and so on.

```{r}
x
y<- matrix(x, nrow = 4)
y

y + 2


z<- array_reshape(x, dim = c(2,3,4)) 
z
z + 2



```


```{r}
array6<- replicate(3, matrix(rnorm(24),ncol=4), simplify=FALSE) %>%
        simplify2array()
        
dim(array6)
array6

## turn all negative to 0, the element-wise relu
pmax(array6, 0)

```

### tensor dot

for 2D tensors, it is like matrix multiplication, the number of columns for the first matrix
has to be the same as the number of rows for the second matrix

```{r}
mat2<- sample(c(0,1), size= 24, replace = TRUE) %>% matrix(ncol = 4)
mat2

mat3<- sample(c(0,1), size= 28, replace = TRUE) %>% matrix(nrow = 4)
mat3

mat2 %*% mat3
```

6x4 mat1 dot 4x7 mat2 == 6x7 matrix.

You can take dot product of higher-dimensional tensors following the same rules for shape compatibility. 


**In my next blog post, I will show you how to use TCR sequences to predict cancer vs healthy samples.** Stay tuned!



