---
title: 'How to separate a comma delimited string into multiple lines in R and python '
author: Ming Tommy Tang
date: '2024-05-09'
slug: how-to-separate-a-comma-delimited-string-into-multiple-lines-in-r-and-python
categories:
  - bioinformatics
  - rstats
  - python
tags:
  - python
  - rstats
header:
  caption: ''
  image: ''
editor_options: 
  chunk_output_type: console
---

### The problem

```{r, include=FALSE}
# https://bookdown.org/yihui/rmarkdown-cookbook/cache-lazy.html
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message= FALSE, warning = FALSE,
  cache = FALSE, cache.lazy= FALSE
)
```


```{r}
df<- data.frame(id = c(1,2,3), value = c('x,y', 'z,w', 'a'))
df
```

we want to put x,y in the first row into two rows:

1, x

1, y

and put z,w into two rows too.

### solution with R

There is a neat function `separate_rows` that does exactly this in `tidyr` package:
```{r}
library(tidyr)
library(dplyr)

df %>%
        tidyr::separate_rows(value, sep=",")
        
```

We can also convert it back using some purrr tricks.

```{r}
library(tidyr)
library(purrr)

df_nest<- df %>%
        tidyr::separate_rows(value, sep=",") %>%
        group_by(id) %>%
        nest()

df_nest

df_nest$data[[1]]

# convert it back 
df_nest %>%
        mutate(new_data = map_chr(data, ~paste(.x$value, collapse = ",")))
```

### solution with python

Note, how to use python in Rmarkdown, read https://rstudio.github.io/reticulate/articles/versions.html

```{r}
library(reticulate)
use_condaenv("r-reticulate")
```

```{python}
import pandas as pd

df = pd.DataFrame({'id': [1, 2, 3], 'data': ['x, y', 'z, w', 'a']})

print(df)

# Split text into a list
df['data'] = df['data'].str.split(',')
print(df)

# Convert list into multiple rows
df = df.explode('data')
print(df)
```
we used `explode` function in pandas. Of course, you can do it manually:
read in the dataframe line by line, split the last column to a list using str.split(',') and open an empty file and write the str.splited entries into new lines.

### use case in bioinformatics?

follow https://yulab-smu.top/biomedical-knowledge-mining-book/clusterprofiler-go.html
for gene set enrichment analysis.

```{r}
library(clusterProfiler)
data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

# Entrez gene ID
head(gene)
```

Let's do a go-term over-representation analysis

```{r}
library(org.Hs.eg.db)

ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
head(ego)
```

The last `geneID` column is a list of genes separated by `/`. If I want to 
get a tidy dataframe, I can do

```{r}
ego@result %>% 
        tidyr::separate_rows(geneID, sep="/")
```

Now each gene has its own row!
