---
title: "PCA analysis on scATACseq data"
author: Ming Tommy Tang
date: "2025-02-02"
slug: pca-scATAC
categories:
  - review
tags:
  - bioinformatics
  - scATAC
  - data
  - pca
header:
  caption: ''
  image: ''
editor_options: 
  chunk_output_type: console
---

**To not miss a post like this, sign up for my [newsletter](https://divingintogeneticsandgenomics.ck.page/profile) to learn computational
biology and bioinformatics.**

```{r, include=FALSE}
# https://bookdown.org/yihui/rmarkdown-cookbook/cache-lazy.html
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message= FALSE, warning = FALSE,
  cache = FALSE, cache.lazy= FALSE
)
```

In my last post, I showed you how to use [PCA for bulk RNAseq data](https://divingintogeneticsandgenomics.com/post/pca-tcga/).

Today, let's see how we can use it for scATACseq data.

Download the example dataset from 10x genomics https://support.10xgenomics.com/single-cell-atac/datasets/1.1.0/atac_pbmc_5k_v1

The dataset is 5k Peripheral blood mononuclear cells (PBMCs) from a healthy donor (v1.0).

Download the `atac_pbmc_5k_v1_filtered_peak_bc_matrix.tar.gz` file and unzip it.
It will give you a folder `filtered_peak_bc_matrix` containing three files:

- barcodes.tsv
- matrix.mtx
- peaks.bed

```{r}
library(Seurat)
library(Matrix)
library(readr)
library(dplyr)
mat<- readMM("~/blog_data/filtered_peak_bc_matrix/matrix.mtx")

# this is a sparse matrix 
mat[1:5, 1:5]

peaks<- read_tsv("~/blog_data/filtered_peak_bc_matrix/peaks.bed", col_names = FALSE)
head(peaks)

peaks<- peaks %>%
        mutate(peak = paste(X1,X2,X3, sep="_"))

barcodes<- read_tsv("~/blog_data/filtered_peak_bc_matrix/barcodes.tsv", col_names = FALSE)
head(barcodes)
```

set the rownames and colnames for the count matrix 
```{r}
rownames(mat)<- peaks$peak
colnames(mat)<- barcodes$X1

mat[1:5, 1:5]
```

check the range of the matrix 
```{r}
range(mat)

dim(mat)
```

It has 47520 peaks and 4621 cells.

### PCA for the raw counts 

Let’s calculate the PCA from scratch with `irlba` for big matrix. Use built-in `svd` if the matrix is small or use `prcomp`. They are all for the same purpose: calculate PCA. 

```{r}
mat.pca<- irlba::irlba(t(mat))

cell_embeddings<- mat.pca$u %*% diag(mat.pca$d)

dim(cell_embeddings)
cell_embeddings[1:5, 1:5]

plot(cell_embeddings[, 1], cell_embeddings[,2])
```

The PCA looks a little strange. 

### PCA after TF-IDF transformation

**Latent Semantic Indexing** or LSI method and used a transformation of the binarized scATAC count matrix called ’TF-IDF` (term frequency–inverse document frequency) which is used in text mining followed by PCA.


**What is TF-IDF Transformation and Why is it Useful in scATAC-seq Analysis?**  

**TF-IDF (Term Frequency-Inverse Document Frequency)** is a numerical transformation originally from **text mining**. In **scATAC-seq**, it helps normalize sparse and binary data to highlight **important peaks** for downstream analysis.  

In text mining, **TF-IDF** measures how important a word (term) is in a document relative to all other documents. It consists of:  

🔹 **TF (Term Frequency)** → How often a term appears in a document.  
🔹 **IDF (Inverse Document Frequency)** → Downweights commonly occurring terms across all documents.  

Mathematically:  
\[
TF = \frac{\text{Raw count of term in a document}}{\text{Total terms in the document}}
\]
\[
IDF = \log \left( \frac{\text{Total number of documents}}{\text{Number of documents containing the term}} \right)
\]
\[
TF-IDF = TF \times IDF
\]

### Why is TF-IDF Used in scATAC-seq?

Unlike scRNA-seq (which has more **continuous expression values**), scATAC-seq data is largely **binary** (1 = accessible, 0 = inaccessible). Raw counts are highly sparse and noisy.  

**Problem**: Some peaks (e.g., promoters of housekeeping genes) are open in many cells and dominate the dataset.

**Solution**: TF-IDF downweights **ubiquitous peaks** and emphasizes **cell-type-specific** regulatory regions. 

Let's implement it in R:

```{r}
# make a copy of the original matrix 
mat2<- mat

# binarize the matrix 
mat2@x[mat2@x >0]<- 1 

# a custom function to perform TF-IDF transformation 
TF.IDF.custom <- function(data, verbose = TRUE) {
  if (class(x = data) == "data.frame") {
    data <- as.matrix(x = data)
  }
  if (class(x = data) != "dgCMatrix") {
    data <- as(object = data, Class = "dgCMatrix")
  }
  if (verbose) {
    message("Performing TF-IDF normalization")
  }
  npeaks <- Matrix::colSums(x = data)
  tf <- t(x = t(x = data) / npeaks)
  # log transformation
  idf <- log(1+ ncol(x = data) / Matrix::rowSums(x = data))
  norm.data <- Diagonal(n = length(x = idf), x = idf) %*% tf
  norm.data[which(x = is.na(x = norm.data))] <- 0
  return(norm.data)
}
```

### PCA on the TF-IDF transformed matrix 

```{r}
mat2<- TF.IDF.custom(mat2)

# what's the range after transformation?
range(mat2)

mat2[1:5, 1:5]

library(irlba)
set.seed(123)
mat.lsi<- irlba::irlba(t(mat2), 50)

# peak_loadings <- mat.lsi$v  

cell_embeddings <- mat.lsi$u %*% diag(mat.lsi$d)  # Cell embeddings (5k cells in rows)
dim(cell_embeddings)


cell_embeddings[1:5, 1:5]

plot(cell_embeddings[, 1], cell_embeddings[, 2])
```

You can see that PCA on the TF-IDF transformed matrix reveals more distinct clusters. 

### The first PC is correlated with sequencing depth

```{r}
seq_depth<- colSums(mat2)

abs(cor(seq_depth, cell_embeddings[, 1]))
abs(cor(seq_depth, cell_embeddings[, 2]))

abs(cor(seq_depth, cell_embeddings[, 3]))
```

The first two PCs are highly correlated with sequencing depth. 
That's why for downstream analysis, the first PC is usually discarded.


Alternatively, you can use `Signac` package to do the same.

```{r eval=FALSE}
library(Signac)

# Apply TF-IDF transformation to scATAC-seq data
atac <- RunTFIDF(atac)
```
  
### Conclusions

- LSI is just PCA on the binarized scATACseq count matrix after TF-IDF transformed.

- In scATAC-seq, TF-IDF transformation helps normalize sparse and binary data to highlight important peaks for downstream analysis.

- Read my old blog post [clustering scATACseq data: the TF-IDF way](https://divingintogeneticsandgenomics.com/post/clustering-scatacseq-data-the-tf-idf-way/)

- Bonus: [How to read PCA plots](https://www.nxn.se/p/how-to-read-pca-plots)
