---
title: "PCA analysis on TCGA bulk RNAseq data"
author: Ming Tommy Tang
date: "2025-01-29"
slug: pca-tcga
categories:
  - review
tags:
  - bioinformatics
  - TCGA
  - RNAseq
  - data
  - pca
header:
  caption: ''
  image: ''
editor_options: 
  chunk_output_type: console
---

**To not miss a post like this, sign up for my [newsletter](https://divingintogeneticsandgenomics.ck.page/profile) to learn computational
biology and bioinformatics.**

```{r, include=FALSE}
# https://bookdown.org/yihui/rmarkdown-cookbook/cache-lazy.html
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message= FALSE, warning = FALSE,
  cache = FALSE, cache.lazy= FALSE
)
```

### what is PCA?
Principal Component Analysis (PCA) is a mathematical technique used to reduce the dimensionality of large datasets while preserving the most important patterns in the data. 

It transforms the original high-dimensional data into a smaller set of new variables called principal components (PCs), which capture the most variation in the data.


### Key Concepts of PCA:

* Dimensionality Reduction – PCA reduces complex datasets with many features (e.g., thousands of genes) into a few key components, making analysis easier.

* Variance Maximization – The first principal component (PC1) captures the most variation in the data, the second (PC2) captures the second most, and so on.

* Orthogonality – Principal components are uncorrelated (perpendicular to each other), ensuring that each captures a unique aspect of the data.


Please read my previous posts on PCA:

- [PCA in action](https://divingintogeneticsandgenomics.com/post/pca-in-action/) Step by step guide to calculate PCA with R

- [permutation test for PCA components](https://divingintogeneticsandgenomics.com/post/permute-test-for-pca-components/)

- [How PCA projection and cell label transfer work in Seurat](https://divingintogeneticsandgenomics.com/post/pca-projection/)

### PCA on TCGA data

In this example, we will download TCGA LUAD and LUSC RNAseq counts data, and do a PCA.
You will see when we do not normalize the counts matrix with total library size.
The first PC (PC1) is correlated with sequencing depth.

Let's dive in!

### Download TCGA data using TCGAbiolinks

There are many ways to get [TCGA data](https://www.cancer.gov/ccg/research/genome-sequencing/tcga). I like [`TCGAbiolinks`](https://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html) the best as it is very easy.


It takes long time to download the data, so I saved it as and Rdata.
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)
library(here)
```

Download TCGA data using `TCGAbiolinks`.

```{r eval=FALSE}
# Load TCGA data (example for LUAD - Lung Adenocarcinoma)
LUAD_query <- GDCquery(project = "TCGA-LUAD",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts")

GDCdownload(LUAD_query)

# this returns a summarizedExperiment object 
TCGA_LUAD_data <- GDCprepare(LUAD_query)

saveRDS(TCGA_LUAD_data, "~/blog_data/TCGA_LUAD_SummarizedExperiment.rds"))
```

Read in the saved RDS file:
```{r}
TCGA_LUAD_data<- readRDS("~/blog_data/TCGA_LUAD_SummarizedExperiment.rds")

## there are many metadata for each sample
colData(TCGA_LUAD_data) %>% 
        colnames() %>%
        tail()


# CIMP methylation subtypes
colData(TCGA_LUAD_data) %>%
  as.data.frame() %>%
  janitor::tabyl(`paper_CIMP.methylation.signature.`)

# save the raw counts matrix 
TCGA_LUAD_mat<- assay(TCGA_LUAD_data)


TCGA_LUAD_mat[1:5, 1:5]

dim(TCGA_LUAD_mat)
```
We have 600 samples.

You noticed that the gene name is in ENSEMBLE ID.

We will download the TCGA LUSC data in the same way.
```{r eval=FALSE}
LUSC_query <- GDCquery(project = "TCGA-LUSC",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts")

GDCdownload(LUSC_query)

TCGA_LUSC_data <- GDCprepare(LUSC_query)

saveRDS(TCGA_LUSC_data, "~/blog_data/TCGA_LUSC_SummarizedExperiment.rds"))
```

read in the saved RDS 
```{r}
TCGA_LUSC_data<- readRDS("~/blog_data/TCGA_LUSC_SummarizedExperiment.rds")

# different transcription subtypes
colData(TCGA_LUSC_data) %>%
  as.data.frame() %>%
  janitor::tabyl(paper_Expression.Subtype)

TCGA_LUSC_mat<- assay(TCGA_LUSC_data)
```


### convert ENSEMBL id to gene symbols

```{r}
library(org.Hs.eg.db)

# remove the version number in the end of the ENSEMBL id
TCGA_LUAD_genes<- rownames(TCGA_LUAD_mat) %>%
  tibble::enframe() %>%
  mutate(ENSEMBL =stringr::str_replace(value, "\\.[0-9]+", ""))

head(TCGA_LUAD_genes)

# there are duplicated gene symbols for different ENSEMBL id
clusterProfiler::bitr(TCGA_LUAD_genes$ENSEMBL, 
                      fromType = "ENSEMBL",
                      toType = "SYMBOL",
                      OrgDb = org.Hs.eg.db) %>%
        janitor::get_dupes(SYMBOL) %>%
        head()

# just keep one of them (simple solution)
TCGA_LUAD_gene_map<- clusterProfiler::bitr(TCGA_LUAD_genes$ENSEMBL,
                                           fromType = "ENSEMBL",
                                           toType = "SYMBOL",
                                           OrgDb = org.Hs.eg.db) %>%
        distinct(SYMBOL, .keep_all = TRUE) 
       
                       
TCGA_LUAD_gene_map<- TCGA_LUAD_gene_map %>%
  left_join(TCGA_LUAD_genes)

head(TCGA_LUAD_gene_map)
```

Subset the gene expression matrices and replace the rownames to gene symbol.
```{r}
TCGA_LUSC_mat<- TCGA_LUSC_mat[TCGA_LUAD_gene_map$value, ]
row.names(TCGA_LUSC_mat)<- TCGA_LUAD_gene_map$SYMBOL
TCGA_LUSC_mat[1:5, 1:5]


TCGA_LUAD_mat<- TCGA_LUAD_mat[TCGA_LUAD_gene_map$value, ]
row.names(TCGA_LUAD_mat)<- TCGA_LUAD_gene_map$SYMBOL
TCGA_LUAD_mat[1:5, 1:5]

## check the dimentions 
dim(TCGA_LUSC_mat)
dim(TCGA_LUAD_mat)
```


### combine TCGA LUSC and LUAD

PCA plot use top variable genes.

```{r}
# double check the genes are the same
all.equal(rownames(TCGA_LUAD_mat), rownames(TCGA_LUSC_mat))

TCGA_lung_mat<- cbind(TCGA_LUSC_mat, TCGA_LUAD_mat)

TCGA_lung_meta<- data.frame(cancer_type = c(rep( "LUSC", ncol(TCGA_LUSC_mat)), 
                   rep("LUAD", ncol(TCGA_LUAD_mat))))

dim(TCGA_lung_mat)
```


### PCA with the raw counts 

```{r}
library(ggplot2)
library(ggfortify)
# select the top 1000 most variable genes 
TCGA_gene_idx<- order(rowVars(TCGA_lung_mat), decreasing = TRUE)[1:1000]

TCGA_lung_mat_sub <- TCGA_lung_mat[TCGA_gene_idx, ]

TCGA_pca_res <- prcomp(t(TCGA_lung_mat_sub), scale. = TRUE)

autoplot(TCGA_pca_res, data = TCGA_lung_meta , color ="cancer_type") +
  scale_color_manual(values = c("blue", "red")) +
  ggtitle("TCGA NSCLC")

```
LUAD and LUSC samples are not separated in the first PC.

Let's see how sequencing depth is correlated with first PC

read https://divingintogeneticsandgenomics.com/post/pca-in-action/ for more details for PCA.

```{r}
# the PCs 
TCGA_pca_res$x[1:5, 1:5]

seq_depth<- colSums(TCGA_lung_mat)

# calculate the correlation of first PC and the sequencing depth
# the sign of the PCs are arbitrary, so let's get the absolute number
cor(TCGA_pca_res$x[, 1], seq_depth) %>%
        abs()
```

A correlation of `0.98` for PC1!!

```{r}
# PC2 correlation with sequencing depth
cor(TCGA_pca_res$x[, 2], seq_depth) %>%
        abs()
```
PC2 is not correlated with sequencing depth. As we can see PC2 separates the cancer types.

Let's plot it on PCA.
```{r}
TCGA_lung_meta$seq_depth<- seq_depth

autoplot(TCGA_pca_res, data = TCGA_lung_meta , color ="seq_depth") +
        scale_color_viridis_b() +
  ggtitle("TCGA NSCLC")
```

You see the sequencing depth is high to low from left to right in PC1.

### PCA after cpm (counts per million) normalization

```{r}
# convert the raw counts to log2(cpm+1) using edgeR.
TCGA_lung_cpm<- edgeR::cpm(TCGA_lung_mat, log = TRUE, prior.count = 1)

## top 1000 most variable genes
TCGA_gene_idx2<- order(rowVars(TCGA_lung_cpm), decreasing = TRUE)[1:1000]

TCGA_lung_cpm_sub <- TCGA_lung_cpm[TCGA_gene_idx2, ]

TCGA_pca_res2 <- prcomp(t(TCGA_lung_cpm_sub), scale. = TRUE)

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="cancer_type") +
  scale_color_manual(values = c("blue", "red")) +
  ggtitle("TCGA NSCLC")
```

Now we see the first PC separates cancer type: LUSC vs LUAD, which makes sense!
It is interesting to see some of the LUSC samples mix with the LUAD.
It will be interesting to further investigate those samples.

### make a heatmap

My favorite package is `ComplexHeatmap`.

```{r}
set.seed(123)
library(ComplexHeatmap)

TCGA_ha<- HeatmapAnnotation(df = TCGA_lung_meta,
    col = list(
               cancer_type = c("LUSC" = "red", "LUAD" = "blue"))
    )

Heatmap(t(scale(t(TCGA_lung_cpm_sub))),
        name = "TCGA lung RNA",
        show_column_names = FALSE,
        show_row_names = FALSE,
        show_row_dend = FALSE,
        #column_split = TCGA_lung_meta$cancer_type,
        top_annotation = TCGA_ha,
        row_names_gp = gpar(fontsize = 3),
        border = TRUE,
        row_km = 3
        )

```

This shows you how useful PCA is. In my next post, I will show you how a similar 
analysis is done in a single-cell ATACseq dataset. Stay tuned!

**To not miss a post like this, sign up for my [newsletter](https://divingintogeneticsandgenomics.ck.page/profile) to learn computational
biology and bioinformatics.**

Happy Learning!

Tommy 