---
title: "PCA analysis on TCGA bulk RNAseq data continued"
author: Ming Tommy Tang
date: "2025-02-07"
slug: pca-tcga2
categories:
  - review
tags:
  - bioinformatics
  - TCGA
  - RNAseq
  - data
  - pca
header:
  caption: ''
  image: ''
editor_options: 
  chunk_output_type: console
---

**To not miss a post like this, sign up for my [newsletter](https://divingintogeneticsandgenomics.ck.page/profile) to learn computational
biology and bioinformatics.**

```{r, include=FALSE}
# https://bookdown.org/yihui/rmarkdown-cookbook/cache-lazy.html
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message= FALSE, warning = FALSE,
  cache = FALSE, cache.lazy= FALSE
)
```
In my [last blog post](https://divingintogeneticsandgenomics.com/post/pca-tcga/), I showed you how to download TCGA RNAseq count data and 
do PCA and make a heatmap. It is interesting to see some of the LUSC samples mix with the LUAD samples and vice versa.

In this post, we will continue to use PCA to do more Exploratory data analysis (EDA).

Let's further investigate those samples and use the rank matrix derived from the count matrix to do PCA.

### Download and combine the data

The following code is the same as my last post.

It takes long time to download the data, so I saved it as and Rdata.
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)
library(here)
```

Download TCGA data using `TCGAbiolinks`.

```{r eval=FALSE}
# Load TCGA data (example for LUAD - Lung Adenocarcinoma)
LUAD_query <- GDCquery(project = "TCGA-LUAD",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts")

GDCdownload(LUAD_query)

# this returns a summarizedExperiment object 
TCGA_LUAD_data <- GDCprepare(LUAD_query)

saveRDS(TCGA_LUAD_data, "~/blog_data/TCGA_LUAD_SummarizedExperiment.rds"))
```

Read in the saved RDS file:
```{r}
TCGA_LUAD_data<- readRDS("~/blog_data/TCGA_LUAD_SummarizedExperiment.rds")
TCGA_LUAD_mat<- assay(TCGA_LUAD_data)
dim(TCGA_LUAD_mat)
```
We have 600 samples.

You noticed that the gene name is in ENSEMBLE ID.

We will download the TCGA LUSC data in the same way.
```{r eval=FALSE}
LUSC_query <- GDCquery(project = "TCGA-LUSC",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts")

GDCdownload(LUSC_query)

TCGA_LUSC_data <- GDCprepare(LUSC_query)

saveRDS(TCGA_LUSC_data, "~/blog_data/TCGA_LUSC_SummarizedExperiment.rds"))
```

read in the saved RDS 
```{r}
TCGA_LUSC_data<- readRDS("~/blog_data/TCGA_LUSC_SummarizedExperiment.rds")

TCGA_LUSC_mat<- assay(TCGA_LUSC_data)

dim(TCGA_LUSC_mat)
```


### convert ENSEMBL id to gene symbols

```{r}
library(org.Hs.eg.db)

# remove the version number in the end of the ENSEMBL id
TCGA_LUAD_genes<- rownames(TCGA_LUAD_mat) %>%
  tibble::enframe() %>%
  mutate(ENSEMBL =stringr::str_replace(value, "\\.[0-9]+", ""))

head(TCGA_LUAD_genes)

# there are duplicated gene symbols for different ENSEMBL id
clusterProfiler::bitr(TCGA_LUAD_genes$ENSEMBL, 
                      fromType = "ENSEMBL",
                      toType = "SYMBOL",
                      OrgDb = org.Hs.eg.db) %>%
        janitor::get_dupes(SYMBOL) %>%
        head()

# just keep one of them (simple solution)
TCGA_LUAD_gene_map<- clusterProfiler::bitr(TCGA_LUAD_genes$ENSEMBL,
                                           fromType = "ENSEMBL",
                                           toType = "SYMBOL",
                                           OrgDb = org.Hs.eg.db) %>%
        distinct(SYMBOL, .keep_all = TRUE) 
       
                       
TCGA_LUAD_gene_map<- TCGA_LUAD_gene_map %>%
  left_join(TCGA_LUAD_genes)

head(TCGA_LUAD_gene_map)
```

Subset the gene expression matrices and replace the rownames to gene symbol.
```{r}
TCGA_LUSC_mat<- TCGA_LUSC_mat[TCGA_LUAD_gene_map$value, ]
row.names(TCGA_LUSC_mat)<- TCGA_LUAD_gene_map$SYMBOL
TCGA_LUSC_mat[1:5, 1:5]


TCGA_LUAD_mat<- TCGA_LUAD_mat[TCGA_LUAD_gene_map$value, ]
row.names(TCGA_LUAD_mat)<- TCGA_LUAD_gene_map$SYMBOL
TCGA_LUAD_mat[1:5, 1:5]

## check the dimentions 
dim(TCGA_LUSC_mat)
dim(TCGA_LUAD_mat)
```


### combine TCGA LUSC and LUAD

PCA plot use top variable genes.

```{r}
# double check the genes are the same
all.equal(rownames(TCGA_LUAD_mat), rownames(TCGA_LUSC_mat))

TCGA_lung_mat<- cbind(TCGA_LUSC_mat, TCGA_LUAD_mat)

TCGA_lung_meta<- data.frame(cancer_type = c(rep( "LUSC", ncol(TCGA_LUSC_mat)), 
                   rep("LUAD", ncol(TCGA_LUAD_mat))))

rownames(TCGA_lung_meta)<- c(colnames(TCGA_LUAD_mat), colnames(TCGA_LUSC_mat))

head(TCGA_lung_meta)
dim(TCGA_lung_mat)
```

### PCA after cpm (counts per million) normalization

```{r}
library(ggplot2)
library(ggfortify)

# convert the raw counts to log2(cpm+1) using edgeR.
TCGA_lung_cpm<- edgeR::cpm(TCGA_lung_mat, log = TRUE, prior.count = 1)

select_top_var_genes<- function(mat, top_n=1000){
        idx<- order(rowVars(TCGA_lung_cpm), decreasing = TRUE)[1:top_n]
        return(mat[idx,])
}
## top 1000 most variable genes

TCGA_lung_cpm_sub <- select_top_var_genes(TCGA_lung_cpm, top_n = 1000)

TCGA_pca_res2 <- prcomp(t(TCGA_lung_cpm_sub), scale. = TRUE)

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="cancer_type") +
  scale_color_manual(values = c("blue", "red")) +
  ggtitle("TCGA NSCLC")
```

Now we see the first PC separates cancer type: LUSC vs LUAD, which makes sense!
It is interesting to see some of the LUSC samples mix with the LUAD samples and vice versa.

Let's further investigate those samples.

### LUAD and LUSC markers

LUAD originates in the alveolar epithelial cells and is often associated with EGFR mutations, particularly in non-smokers.

NKX2-1 (TTF-1)	
Lung epithelial differentiation marker. 
Highly expressed in LUAD but absent in LUSC (TCGA, IHC studies). 

NAPSA (Napsin A). 
Aspartic protease in surfactant protein processing. 
High specificity for LUAD (IHC studies). 

```{r}
TCGA_lung_meta$NAPSA<- TCGA_lung_cpm["NAPSA", ]
TCGA_lung_meta$TFF1<- TCGA_lung_cpm["TFF1", ]

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="TFF1") +
        scale_color_viridis_b() +
        facet_wrap(~ cancer_type) + 
        ggtitle("TCGA NSCLC")
```

Let's cap the TFF1 max value at 6

```{r}

quantile(TCGA_lung_meta$TFF1, c(0.1,0.2,0.5,0.8,0.9, 0.951))

# Cap the maximum value at 6
TCGA_lung_meta$TFF1 <- pmin(TCGA_lung_meta$TFF1, 6)

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="TFF1") +
        scale_color_viridis_b() +
        facet_wrap(~ cancer_type) + 
        ggtitle("TCGA NSCLC")
```
TFF1 mRNA is not high across the LUAD samples. TFF1 protein maybe a better marker?

```{r}
quantile(TCGA_lung_meta$NAPSA, c(0.1,0.2,0.5,0.8,0.9, 0.951))

# Cap the maximum value at 8
TCGA_lung_meta$NAPSA <- pmin(TCGA_lung_meta$NAPSA, 8)

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="NAPSA") +
        scale_color_viridis_b() +
        facet_wrap(~ cancer_type) + 
        ggtitle("TCGA NSCLC")

```

NAPSA level looks pretty high in most of the LUAD samples. The LUAD samples at PC1 > 0 have lower NAPSA levels. Also, the LUSC samples at PC1 < 0 have high NAPSA levels too.

Are those samples mislabeled?

Let's check several LUSC markers.

LUSC arises from the bronchial epithelium and is strongly associated with smoking.

TP63 (ΔNp63). 	
Transcription factor crucial for squamous differentiation. 	
High expression in LUSC, absent in LUAD (TCGA, IHC studies). 

KRT5 (Cytokeratin 5). 	
Cytoskeletal protein in basal epithelial cells. 	
Expressed in LUSC, absent in LUAD. 

```{r}
TCGA_lung_meta$TP63<- TCGA_lung_cpm["TP63", ]
TCGA_lung_meta$KRT5<- TCGA_lung_cpm["KRT5", ]

TCGA_lung_meta$TP63 <- pmin(TCGA_lung_meta$TP63, 8)
autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="TP63") +
        scale_color_viridis_b() +
        facet_wrap(~ cancer_type) + 
        ggtitle("TCGA NSCLC") 

autoplot(TCGA_pca_res2, data = TCGA_lung_meta , color ="KRT5") +
        scale_color_viridis_b() +
        facet_wrap(~ cancer_type) + 
        ggtitle("TCGA NSCLC")
```
Both TP63 and KRT5 are very high in most of the LUSC samples except the samples that are located in PC1 < 0. 


### Isolate those samples based on PC1

```{r}
PC1_greater_zero<- TCGA_pca_res2$x[, 1] > 0

PC1_smaller_zero<- TCGA_pca_res2$x[, 1] < 0 

TCGA_lung_meta$PC1_greater_zero<- PC1_greater_zero
TCGA_lung_meta$PC1_smaller_zero<- PC1_smaller_zero

head(TCGA_lung_meta)

### samples that are labeled as LUAD but mixed with LUSC
TCGA_lung_meta %>%
        filter(cancer_type == "LUAD", PC1_greater_zero)
```
One can further investigate those samples (by imaging data or proteomics data) and see if there could be mislabeled.

```{r}
### samples that are labeled as LUSC but mixed with LUAD
TCGA_lung_meta %>%
        filter(cancer_type == "LUSC", PC1_smaller_zero) %>%
        head()
```

### PCA on the rank matrix

There could be certain batch effect that we do not know.

and the batch effect somehow force the LUAD and LUSC mixing in the PCA plot.

Let's try rank the genes per sample by their expression level and use the rank
matrix to do PCA.

```{r}
# Convert to rank matrix (ranking each column separately)
# TPM may be better than CPM 
# rank_matrix <- apply(TCGA_lung_cpm, 2, rank, ties.method = "average")
# faster than the apply

library(matrixStats)
rank_matrix <- colRanks(TCGA_lung_cpm, ties.method = "average")

## colRanks returns sample x gene, transpose it back to gene x sample
rank_matrix<- t(rank_matrix)

hist(rank_matrix)
```

PCA assumes continuous, approximately normal data because it relies on variance computation.
Rank values are discrete and not normally distributed.
However, in practice, PCA can still work on ranked data, though its interpretation changes.

```{r}
rank_matrix_sub<- select_top_var_genes(rank_matrix, top_n = 1000)
hist(rank_matrix_sub)
```
It looks approximately normal.

PCA on this matrix:
```{r}
pca_result_rank<- prcomp(t(rank_matrix_sub), scale. = TRUE)

autoplot(pca_result_rank, data = TCGA_lung_meta , color ="cancer_type") +
  ggtitle("TCGA NSCLC raw rank")

autoplot(pca_result_rank, data = TCGA_lung_meta , color ="cancer_type") +
  facet_wrap(~cancer_type) +
  ggtitle("TCGA NSCLC raw rank")
```
PCA on the raw rank matrix somehow can separate LUAD and LUSC as well!

### PCA on normalized rank matrix 

This maps ranks to a normal distribution using the inverse normal transformation.
```{r}
rank_normalized<- qnorm((rank_matrix - 0.5) / nrow(rank_matrix))
hist(rank_normalized)
```

```{r}
rank_normalized_sub<- select_top_var_genes(rank_normalized, top_n = 1000)
hist(rank_normalized_sub)
```
It looks approximately normal.

PCA on the normalized rank matrix 
```{r}
pca_result_rank<- prcomp(t(rank_normalized_sub), scale. = TRUE)

autoplot(pca_result_rank, data = TCGA_lung_meta , color ="cancer_type") +
  ggtitle("TCGA NSCLC normalized rank")

autoplot(pca_result_rank, data = TCGA_lung_meta , color ="cancer_type") +
  facet_wrap(~cancer_type) +
  ggtitle("TCGA NSCLC normalized rank")
```

The PCA plot looks similar to the one with the raw rank and the cpm normalized counts.
We still see those LUAD/LUSC mixing samples.

### Key takeaways

- Use PCA do do exploratory data analysis.

- Use your biology knowledge (in our case, LUAD and LUSC known markers) to investigate the strange samples. 

- PCA on the raw rank matrix and the normalized rank matrix can separate the LUAD and LUSC samples.

There is a long post for how to use `DESeq2` on large scale dataset. 

https://bsky.app/profile/iansudbery.bsky.social/post/3lgdp62x5bc2q

In my next blog post, I will show you using `DESeq2` vs `wilcox rank sum` test for
differential gene expression analysis.

### Further reading

- [Removing unwanted variation from large-scale RNA sequencing data with PRPS](https://www.nature.com/articles/s41587-022-01440-w)

- The github repo associated with the paper has a lot of tutorials https://github.com/RMolania/TCGA_PanCancer_UnwantedVariation/

- https://htmlpreview.github.io/?https://github.com/RMolania/TCGA_PanCancer_UnwantedVariation/blob/master/Vigettes/TCGA_BRCA_RNAseq_Vignette.html

- Twitter thread by Mike Love on how to analyze large scale RNAseq data https://x.com/mikelove/status/1513468597288452097?s=46

- Follow Mike Love's tutorial here https://github.com/mikelove/preNivolumabOnNivolumab
